|

### 2022-05-17 article/article.Rmd
edit article

This paper incorporates partisanship and mayoral first-move advantage to a
\citet{baron1989bargaining} bargaining model. Mayors make the initial policy
proposal, which consists of investments in public goods provision, rents for
personal gains, and a vector of side-transfers for councilors. If mayors fail
to secure majoritarian support from councilors, a reversal stage occurs, in
which the decision leaves the mayor's office and councilors propose the public
investments. In this mayor-council bargaining process, costs come from the
side-transfers required to secure a majority among councilors. We show that in
a model without partisan concerns, bargaining costs always increase with larger
city councils, which in turn lower public goods provision and rents. However,
partisanship decreases bargaining costs when the chances of electing
government-aligned councilors are sufficiently high. Lower bargaining costs
release funds for mayors to invest in public goods provision. In this case, a
larger city council may add more members to the government coalition, thus
reducing the mayor's bargaining costs and increasing the equilibrium public
goods provision.

@@

We find that larger city councils changed the composition of mayoral coalitions
and the number of appointed bureaucrats in the municipality. The extra city
councilor had a 91 percent chance of belonging to the mayor's pre-electoral
coalition. With a larger coalition, mayors face lower bargaining costs and are
able to boost public investments with fewer constraints. Each additional
legislator also leads to 104 new appointed bureaucrats, and research shows that
political appointees civil servants improve public service provision and
enhance accountability in Brazil \citep{toral2019benefits}.

@@

Second, this paper contributes to the literature on the effects of larger
legislatures in service provision. Following the seminal contribution by
\citet{Weingastetal1981}, the "law of $1/n$" posits that larger legislatures
increase the legislative pressure for providing targeted public goods to local
supporters at the expense of the entire polity \citep{freire2021legislature}.
This question has become more relevant since legislature size has increased
consistently over time \citep{gerring2020population}. The corollary of the "law
of $1/n$" is that increasing legislature size promotes excessive public
spending, regardless of the partisan makeup of the legislature
\citep{GilliganMatsusaka2001}.\footnote{\citet{PrimoSnyder2008} show that
legislature size may result in positive or negative expenditures, depending on
factors such as the degree that local public goods can exclude non-locals or
deadweight costs associated with taxation.} In the US metaphor, this would
explain significant infrastructure spending in smaller districts. However, the
"law of $1/n$" literature rarely mentions the relationship between public goods
provision and welfare. This paper is one of the first to study how the number
of local legislators affects welfare. Our model provides a theoretical
expectation based on the chances of electing a politician aligned with the
mayor. The empirical estimates confirm our theoretical predictions, with the
additional fact that legislature size has a null effect on expenditures in
Brazil.

Third, our article also contributes to the growing literature on municipal
partisan politics. Most papers in this field estimate the effects of the
mayor's party on political outcomes. For instance, \citet{de2016mayoral} and
\citet{gerber2011mayors} find differences in governance when comparing
Democratic and Republican mayors. Specifically about Brazil,
\citet{gouvea2021partisanship} shows that social policy expenditure increase
with left-wing mayors. \citet{frey2021legislature}, using discontinuities in
city council size after the 2009 constitutional amendment that reverted the
thresholds we study here, shows that changes in council size lower the support
for candidates backed by the municipal mayors. Our paper reinforces the view
that partisanship matters locally, even in countries with weak ideological
linkage.

@@

This paper shows that partisanship considerably decreases bargaining costs and
improves citizen welfare. The crucial actors in providing public goods are the
mayors, but they cannot do it alone. Legislatures are a solid check to the
mayor's policy. Bargaining with them in a country plagued by corruption can be
costly \citep{bertholini2017price}. Therefore, an extra ally in the city
council may help ease the bargaining and increase social welfare.

Increasing the council size results in gains in elementary school enrollment
and in lower infant mortality. Education quality remains unchanged, but
increasing enrollment without affecting quality should be interpreted as a
positive outcome. The analysis of the approved legislation shows that the
frequency of laws on local public goods increases in municipalities above the
council size thresholds. Finally, our survey data demonstrate that councilors
and mayors agree about appointing more public servants to the public
administration.

\citet{Ames2001} argues that the Brazilian democracy is prone to deadlocks
because the executive has its hands tied by legislators that demand consistent
provision of particularistic and private goods. Indeed, the most successful
presidents after 1995 had to build multiparty coalitions by negotiating with
parties \citep{limongi1998bases, raile2011executive}. Despite evidence that
side payments cemented these coalitions, many social and economic policies have
since been implemented. For instance, education and health care notably
improved during the Cardoso and Lula governments \citep{acemoglu2012nations}.
Presidents who challenged the coalition system did not succeed. For instance,
former President Rousseff was impeached in her second term. Bolsonaro can be
considered one of the most ineffective presidents in Brazilian history.

Municipalities work similarly in terms of the incentives for
legislative-executive bargaining. The stakes are lower, and the negotiations
involve only one legislative branch. Still, in a system such as Brazil's, with
more than thirty-five registered parties, mayors face substantial challenges to
the support for their agendas. Having extra help in the council can tilt the
bargaining towards cheaper governing costs, thus improving public goods
provision.

However, we do not argue that these benefits result from the absence of
corruption or clientelism. Both the formal model and empirical evidence
provided by \citet{britto2020corruption} show that corruption increases with
legislature size. As the model shows, city councilors receive a side transfer
from mayors to support their agenda. Mayors also divide the resources saved in
the bargain into public goods and rents. The rate for the division depends on
the marginal changes of the rents and public goods, but these two components
receive the extra resources. In this case, the mayor's rents may be primarily
corruption, and the councilor's transfers can be clientelistic, aimed at
improving the councilor's reelection chances.

>>>>

This paper incorporates partisanship and mayoral first-move advantage to a
\citet{baron1989bargaining} bargaining model. Mayors make the initial policy
proposal, which consists of investments in public goods provision, rents for
personal gains, and a vector of side-transfers for councilors. If mayors fail
to secure majoritarian support from councilors, a reversal stage occurs, in
which the decision leaves the mayor's office and councilors propose the public
investments. In this mayor-council bargaining process, costs come from the
side-transfers required to secure a majority among councilors. We show that in
a model without partisan concerns, bargaining costs always increase with larger
city councils, which in turn lower public goods provision and rents. However,
partisanship decreases bargaining costs when the chances of electing
government-aligned councilors are sufficiently high. Lower bargaining costs
release funds for mayors to invest in public goods. In this case, a larger city
council may add more members to the government coalition, thus reducing the
mayor's bargaining costs and increasing the equilibrium public goods provision.

@@

We find that larger city councils changed the composition of mayoral coalitions
and the number of appointed bureaucrats in the municipality. The extra city
councilor had a 91 percent chance of belonging to the mayor's pre-electoral
coalition. With a larger coalition, mayors face lower bargaining costs and are
able to boost public investments with fewer constraints. Each additional
legislator also leads to 104 new appointed bureaucrats, and research shows that
political appointees improve public service provision and enhance
accountability in Brazil \citep{toral2019benefits}.

@@

Second, this paper contributes to the literature on the effects of larger
legislatures in service provision. Following the seminal contribution by
\citet{Weingastetal1981}, the "law of $1/n$" posits that larger legislatures
increase the legislative pressure for providing targeted public goods to local
supporters at the expense of the entire polity \citep{freire2021legislature}.
This question has become more relevant since legislature size has increased
consistently over time \citep{gerring2020population}. The corollary of the "law
of $1/n$" is that increasing legislature size promotes excessive public
spending, regardless of the partisan makeup of the legislature
\citep{GilliganMatsusaka2001}.\footnote{\citet{PrimoSnyder2008} show that
legislature size may result in positive or negative expenditures, depending on
factors such as the degree that local public goods can exclude non-locals or
deadweight costs associated with taxation.} In the US metaphor, this would
explain significant infrastructure spending in smaller districts. However, the
literature on the "law of $1/n$" rarely mentions the relationship between
public goods provision and welfare. This paper is one of the first to study how
the number of local legislators affects citizen well-being. Our model provides
a theoretical expectation based on the chances of electing a politician aligned
with the mayor. The empirical estimates confirm our theoretical predictions,
with the additional fact that legislature size has a null effect on
expenditures in Brazil.

Third, our article also contributes to the growing literature on municipal
partisan politics. Most papers in this field estimate the effects of the
mayor's party on political outcomes. For instance, \citet{de2016mayoral} and
\citet{gerber2011mayors} find differences in governance when comparing
Democratic and Republican mayors. Specifically about Brazil,
\citet{gouvea2021partisanship} show that social policy expenditure increases
with left-wing mayors. \citet{frey2021legislature}, using discontinuities in
city council size after the 2009 constitutional amendment that reverted the
thresholds we study here, shows that changes in council size lower the support
for candidates backed by the municipal mayors. Our paper reinforces the view
that partisanship matters locally, even in countries with weak ideological
linkage.

@@

This paper shows that partisanship considerably decreases bargaining costs and
improves citizen welfare. The crucial actors in providing public goods are the
mayors, but they cannot do it alone. Legislatures are a solid check to the
mayor's agenda. Bargaining with them in a country plagued by corruption can be
costly \citep{bertholini2017price}. Therefore, an extra ally in the city
council may help ease the bargaining and increase social welfare.

Increasing the council size results in gains in elementary school enrollment
and in lower infant mortality. Education quality remains unchanged, but
increasing enrollment without affecting quality should be interpreted as a
positive outcome. The analysis of the approved legislation shows that the
frequency of laws on local public goods increases in municipalities above the
council size thresholds. Finally, our survey data demonstrate that councilors
and mayors agree about appointing more civil servants to the public
administration.

\citet{Ames2001} argues that the Brazilian democracy is prone to deadlocks
because the executive has its hands tied by legislators who demand consistent
provision of private goods. Indeed, the most successful presidents after 1995
had to build extensive coalitions by negotiating with several parties
\citep{limongi1998bases, raile2011executive}. Despite evidence that side
payments cemented these coalitions, many social and economic policies have
since been implemented. For instance, education and health care notably
improved during the Cardoso and Lula administrations
\citep{acemoglu2012nations}. Presidents who challenged the coalition system did
not succeed. For instance, former President Rousseff was impeached in her
second term. Bolsonaro can be considered one of the most ineffective presidents
in Brazilian history.

Municipalities work similarly in terms of the incentives for
legislative-executive bargaining. The stakes are lower, and the negotiations
involve only one legislative branch. Still, in a system such as Brazil's, with
more than thirty-five registered parties, mayors face substantial challenges to
approve their political agendas. Having extra help in the council can tilt the
bargaining towards cheaper governing costs, thus improving public goods
provision.

However, we do not argue that these benefits result from the absence of
corruption or clientelism. Both the formal model and empirical evidence
provided by \citet{britto2020corruption} show that corruption increases with
legislature size. As the model shows, city councilors may receive a side
transfer from the mayor to support her agenda. Mayors also divide the resources
saved in the bargain into public goods and rents. The rate for the division
depends on the marginal changes of the rents and public goods, but these two
components receive the extra resources. In this case, the mayor's rents may be
primarily corruption, and the councilor's transfers can be clientelistic, aimed
at improving the councilor's reelection chances.
